---
title: "Analyzing NYC Payroll Data for Fiscal Year 2025"
author: Melissa Mugisha Kabalisa
format: html
---

## About The Data

The data shows annual salary and job details for each New York City employee from 2014 to 2025. It is collected by the NYC Office of Payroll Administration (OPA). However the data was filtered to only data in the five boroughs of NYC (Bronx, Brooklyn, Manhattan, Queens, Staten Island) and from the fiscal year 2025. Each row represents a city employee salary per fiscal year.

The original dataset can be found [here](https://data.cityofnewyork.us/City-Government/Citywide-Payroll-Data-Fiscal-Year-/k397-673e/data_preview).

## Data Definition

-   `Fiscal Year`: The data was wrangled to have only 2025 data.
-   `Payroll Number`
-   `Agency Name`: The Payroll agency that the employee works for
-   `Last Name`: Last name of employee
-   `First Name`: First name of employee
-   `Mid Init`: Middle initial of employee
-   `Agency Start Date`: Date which employee began working for their current agency
-   `Work Location Borough`: Borough of employee's primary work location. Filtered to include only NYC boroughs (Bronx, Brooklyn, Manhattan, Queens, Staten Island)
-   `Title Description`: Civil service title description of the employee
-   `Leave Status as of June 30`: Status of employee as of the close of the relevant fiscal year: Active, Ceased, or On Leave
-   `Base Salary`: Base Salary assigned to the employee
-   `Pay Basis`: Lists whether the employee is paid on an hourly, per diem or annual basis
-   `Regular Hours`: Number of regular hours employee worked in the fiscal year
-   `Regular Gross Paid`: The amount paid to the employee for base salary during the fiscal year
-   `OT Hours`: Overtime Hours worked by employee in the fiscal year
-   `Total OT Paid`: Total overtime pay paid to the employee in the fiscal year
-   `Total Other Pay`: Includes any compensation in addition to gross salary and overtime pay, ie Differentials, lump sums, uniform allowance, meal allowance, retroactive pay increases, settlement amounts, and bonus pay, if applicable.

### Bias in the Data

The borough of Staten Island is not at all represented in the data, and while it is the least populated borough, it still has municipal workers who are important to the data visualizations and interpretations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(DescTools)
library(readr)
library(readxl)
library(RColorBrewer)
library(knitr)

nyc_payroll_data_2025 <- read_csv("Citywide_Payroll_data_(Fiscal_Year)_20260215.csv")

nyc_payroll_data_2025 <- nyc_payroll_data_2025 |>
  mutate(`Regular Gross Paid` = parse_number(`Regular Gross Paid`)) |>
  mutate(`Base Salary` = parse_number(`Base Salary`)) |>
  mutate(`Total OT Paid` = parse_number(`Total OT Paid`)) |>
  mutate(`Total Other Pay` = parse_number(`Total Other Pay`)) 
```

```{r results='asis', echo=FALSE}
cat('
<div style="margin-bottom: 20px;">
  <button id="toggleButton" onclick="toggleHighlight()" style="padding: 5px 10px; background-color: #0d4166; color: white; border: none; border-radius: 5px; cursor: pointer;">
    Toggle Table Highlight
  </button>
</div>')
```

## Summary statistics

### Quantitative variables

#### Regular Hours Worked

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025,
  aes(x = `Regular Hours`)
) + 
  geom_histogram(bins = 15, fill = "#0d4166", color = "lightgrey") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Regular Hours Worked",
    x = "Regular Hours",
    y = "Frequency (Number of Employees)"
  )
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
regularhours_summary <- nyc_payroll_data_2025 |>
  summarize(
    Mean = mean(`Regular Hours`, na.rm=TRUE),
    Median = median(`Regular Hours`, na.rm=TRUE),
    SD = sd(`Regular Hours`)
  )
regularhours_summary <- kable(regularhours_summary)
regularhours_summary
```

#### Total OT Paid

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025,
  aes(x = `Total OT Paid`)
) +
  geom_histogram(fill = "#0d4166", color = "white", bins = 45) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Overtime Pay",
    x = "Total OT Paid ($)",
    y = "Frequency"
  )
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
basesalary_summary <- nyc_payroll_data_2025 |>
  summarize(
    Mean = mean(`Total OT Paid`, na.rm = TRUE),
    Median = median(`Total OT Paid`),
    SD = sd(`Total OT Paid`),
    Mode = DescTools::Mode(`Total OT Paid`)
  )
 basesalary_summary <- kable(basesalary_summary)
 basesalary_summary
```

### Ordinal variables

#### Pay Basis

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025,
  aes(
    x = `Pay Basis`
  )
) +
  geom_bar(fill = "#0d4166", color = "lightgrey") +
  labs(
    y = "Number of City Workers"
  ) +
  scale_y_continuous(labels = scales::comma)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pay_basis_stats <- nyc_payroll_data_2025 |>
  group_by(`Pay Basis`) |>
  summarise(
    Frequency = n()
  ) |>
  mutate(
    Percentage = (Frequency / sum(Frequency)) * 100,
    `Cumulative %` = cumsum(Percentage)
  ) |>
  arrange(desc(Frequency))

pay_basis_mode <- DescTools::Mode(nyc_payroll_data_2025$`Pay Basis`, na.rm = TRUE)

pay_basis_stats <- kable(pay_basis_stats)
pay_basis_stats
```

The Mode for Pay Basis is `r pay_basis_mode`.

### Categorical variables

#### Work Location Borough

```{r echo=FALSE, message=FALSE, warning=FALSE}
nyc_payroll_data_2025 <- nyc_payroll_data_2025 |>
  filter(`Work Location Borough` %in% c("BROOKLYN", "BRONX", "MANHATTAN", "QUEENS", "STATEN ISLAND"
  ))

ggplot(
  nyc_payroll_data_2025, 
  aes(
    x = `Work Location Borough`
  )
  ) +
  geom_bar(fill = "#0d4166", color = "lightgrey") +
  labs(
    y = "Number of City Workers"
  ) +
  scale_y_continuous(labels = scales::comma)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
worklocation_summary <- nyc_payroll_data_2025 |>
  count(`Work Location Borough`, name = "Count") |>
  mutate(
    Proportion = Count / sum(Count),
    is_Mode = `Work Location Borough` %in% DescTools::Mode(nyc_payroll_data_2025$`Work Location Borough`, na.rm = TRUE)
  ) |>
  arrange(desc(Proportion))

 worklocation_summary <- kable(worklocation_summary)
```

#### Leave Status as of June 30

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025, 
  aes(
    x = `Leave Status as of June 30`
  )
  ) +
  geom_bar(fill = "#0d4166", color = "lightgrey") +
  labs(
    y = "Number of City Workers"
  ) +
  scale_y_continuous(labels = scales::comma)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
leavestatus_summary <- nyc_payroll_data_2025 |>
  count(`Leave Status as of June 30`, name = "Count") |>
  mutate(
    Proportion = Count / sum(Count),
    is_Mode = `Leave Status as of June 30` %in% DescTools::Mode(nyc_payroll_data_2025$`Leave Status as of June 30`, na.rm = TRUE)
  ) |>
  arrange(desc(Proportion))

 leavestatus_summary <- kable(leavestatus_summary)
 leavestatus_summary
```

#### Categorical Variable: Top 10 NYC Agencies by Employee Count

```{r echo=FALSE, message=FALSE, warning=FALSE}
agency_summary <- nyc_payroll_data_2025 |>
  count(`Agency Name`, name = "Employee_Count") |>
  mutate(Proportion = Employee_Count / sum(Employee_Count)) |>
  slice_max(order_by = Employee_Count, n = 10) |>
  arrange(desc(Employee_Count))

unique_agencies <- length(unique(nyc_payroll_data_2025$`Agency Name`))

agency_summary <- kable(agency_summary)
agency_summary
```

Total number of unique agencies represented: `r unique_agencies`

## Potential relationships

### Pay Basis and Base Salary

#### Testable Hypotheses

Null hypothesis: There is no difference in base salary across pay bases.

Alternative hypothesis: Some pay bases receive higher base salaries than others.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025, 
  aes(
    x = `Pay Basis`,
    y = `Base Salary`
  )
  ) +
  geom_boxplot(color="#0d4166") +
  scale_y_continuous(labels = scales::comma)
```

The boxplot indicates a significant disparity in pay scales based on the Pay Basis. Workers paid per Annum have a much higher and more concentrated median salary compared to Hourly or Per Diem workers. The Prorated Annual category shows the most significant variance/skewing, and underlying reasons for this variation cannot be determined from this boxplot alone.

### Regular Hours and Work Location Borough

#### Testable Hypotheses 

Null hypothesis: There is no difference/variability in regular hours worked across NYC boroughs.

Alternative hypothesis: Some NYC borough(s) display more regular hours worked than others. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025, 
  aes(
    y = `Regular Hours`,
    x = `Work Location Borough`
  )
  ) +
  geom_boxplot(color="#0d4166") +
  scale_y_continuous(labels = scales::comma)
```

While the median regular hours appear similar across boroughs, Manhattan shows a wider spread and more extreme values, indicating greater variability in hours worked. However, the plot does not provide enough information to infer why the differences in median hours.

### Regular Gross Paid and Leave Status as of June 30

#### Testable Hypotheses 

Null hypothesis: There is no difference in gross pay across leave status.

Alternative hypothesis: Employees with certain leave statuses receive a higher gross pay than others. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  nyc_payroll_data_2025, 
  aes(
    y = `Regular Gross Paid`,
    x = `Leave Status as of June 30`
  ) 
  ) +
  geom_boxplot(color="#0d4166") +
  scale_y_continuous(labels = scales::comma)
```

Employees with Active or On Leave status tend to have higher regular gross pay, while those marked ‘Ceased’ have lower values. This pattern may show differences in time worked during the fiscal year, but this cannot be confirmed without additional data.